---
title: "Potential locations for microalgae-based carbon capture facilities in NYS"
author: Abby Stressinger
subtitle: Final Project for GEO511:Spatial Science
date: today
date-format: long
output: 
  html_document:
  theme: cosmo
  code_folding: TRUE
---

# Introduction

Microalgae cultivation for carbon dioxide removal is a potential solution for the climate crisis. Through photosynthesis, microalgae convert carbon dioxide into biomass at rates 10-50x more efficient than land plants. Developing technologies that utilize microalgae for carbon dioxide removal follow two main systems: open and closed systems. Closed systems are those in which inputs (sunlight, nutrients, CO2, water) and growth conditions (temperature, pH, salinity, flow/mixing) are controlled. Assessing optimal locations for closed systems is helpful to determine the long-term scalability and market potential of this technology. To determine optimal locations, multiple factors should be analyzed: biotic (sunlight, air and water temperature) and abiotic factors (carbon sources, land availability, water availability, and economic feasibility). 

For this project, I will assess **potential locations for microalgae-based carbon capture** based on location of carbon point sources and water availability. These factors greatly influence the potential for success of algal biorefineries, since carbon and water are the main controllable  inputs for algal growth. 

I will investigate the following **research questions:**

1. What is the distribution of wastewater facilities across NYS? How does will this compare to carbon sources?

2. What is the distribution of point carbon sources across NYS? How does this compare to across the US? How does this compare to distribution of "high-purity" carbon sources?

3. Based on carbon and water availability, which locations across NYS are best suited for microalgae-based carbon capture?


# Materials and methods

## Data Collection

#### Wastewater Facilites
Data on water availability was collected from wastewater facility data from <https://data.ny.gov>. This data provides information on wastewater treatment plants across NYS. This specific dataset can be downloaded here: <https://data.ny.gov/Energy-Environment/Wastewater-Treatment-Plants/2v6p-juki/about_data>.

#### Carbon Point Sources
Data on carbon sources were obtained from the Environmental Protection Agency 2018 Greenhouse Gas Reporting Program (GHGRP) dataset. This data was filtered by high purity source: Ammonia production, steam methane reforming, and extended to petroleum and natural gas systems (medium purity) for the purpose of applying to NYS. This data is available here: <https://www.epa.gov/ghgreporting>

## Download and Clean Data

Load packages and libraries

```{r setup, message = FALSE, warning = FALSE, echo = FALSE}
#install.packages("sf")
#install.packages("tibble")
#install.packages("dplyr")
#install.packages("leaflet")
#install.packages("ggplot2")
#install.packages("rnaturalearth")
#install.packages("usmap")
#install.packages("mapdata")
#install.packages("maps")
#install.packages("viridis")
#install.packages("spData")

library(sf)
library(tibble)
library(dplyr)
library(leaflet)
library(ggplot2)
library(rnaturalearth)
library(usmap)
library(mapdata)
library(maps)
library(viridis)
library(spData)
```

#### Load in and Clean Wastewater Data

```{r data}
#from https://data.ny.gov/Energy-Environment/Wastewater-Treatment-Plants/2v6p-juki/about_data
data_water <- read.csv("data/Wastewater_Treatment_Plants_20241205.csv")
data_water_NY_sf <- data_water %>%
  filter(!is.na(Longitude) & !is.na(Latitude)) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)

#Load in US country and border data
      #Load in US States data
      data(world)
      data(us_states)
      US_outline <- world %>%
        filter(name_long == "United States") %>%
        st_transform(crs = 4326)
      
      New_York <- us_states %>%
        filter(NAME == "New York") %>%
        st_transform(crs = 4326)
```

Filter and Plot largest facilities:

```{r}
#Filter to largest facilities
  #filter to plants about the mean flow
  summary(data_water_NY_sf$Average.Design.Hydraulic.Flow)

  data_water_NY_above_mean <-
    filter(data_water_NY_sf, Average.Design.Hydraulic.Flow > 13.89)

  #Plot largest facilities
  #ggplot(data_water_NY_above_mean) + geom_sf(aes(color = `Plant.Type`))
  
   ggplot(data_water_NY_above_mean) +
        geom_sf(aes(color = `Plant.Type`)) +
      geom_sf(data = New_York, fill = NA) +
     labs(title = "Large Wastewater Facilites in NYS (2023)",
          subtitle = "Facilities with above average hydraulic flow")
          
```

#### Load in and Clean Carbon Data

Clean and filter US data
```{r}
data <- read.csv("data/ghgp_data_2023_NY.csv")

#Clean up dataset for Entire US
  #change column names and delete NA rows
  colnames <- data[3,]
  data2 <- data[-(1:3),]
  colnames(data2) = colnames

  #Filter out empty rows, and empty long/lat rows
  data_US <- data2[1:6470, ] %>%
   filter(!is.na(Longitude) & !is.na(Latitude)) #too many points

    #check for missing values
    #MISSING_COORD <- is.na(data3$Longitude)
                 #is.na(data3$Latitude)
      #sum(MISSING_COORD$TRUE) #0

  #Transform to shapefile with geometry
  data_US_sf <- st_as_sf(data_US, coords = c("Longitude", "Latitude"), crs = 4326)

  #filter by emissions type (pure carbon sources)
  data_US_pure_carbon <- data_US_sf %>%
   filter(`Industry Type (subparts)`== "P" | 
            `Industry Type (subparts)`== "G" | 
            `Industry Type (subparts)`== "C,W-NGTC")
      #no ammonia plants in US?
```

Clean and filter NY Data
```{r}
#Filter dataset to only NY
  #convert data to a sf object
  data_NY <- filter(data2, State == "NY")

  #str(data_NY_sf) #check that is shapefile, coordinate system, bounding box
  data_NY_sf <- st_as_sf(data_NY, coords=c("Longitude", "Latitude"), crs = 4326)

    #filter NY to high purity carbon sources
  data_NY_pure_carbon <- data_NY_sf %>%
    filter(`Industry Type (subparts)`== "P" | 
             `Industry Type (subparts)`== "G" | 
             `Industry Type (subparts)`== "C,W-NGTC")

```

Plot all US Carbon Sources

```{r}

```

Plot high-purity US Carbon Sources

```{r}
ggplot() +
  geom_sf(data = US_outline, fill = NA) +
  geom_sf(data = data_US_pure_carbon, 
          aes(color = "Industry Type (subparts)", 
              size = "Total reported direct emissions")) +
  guides(size = FALSE) +
  labs(title = "High-purity carbon sources across US")
```


# Results

The plots below visualize the locations of wastewater and carbon source data for NYS. In the last plot, we see multiple areas suitable for microalgae-cultivation, due to overlap of these locations. These locations are specified in the Conclusion.

#### Map Carbon Emissions and Wastewater Data

Map all of carbon point sources in NY

```{r}
#Map all of carbon emissions in NY
ggplot() +
  geom_sf(data = New_York, fill = NA) +
  geom_sf(data = data_NY_sf, aes(color = `Industry Type (sectors)`,
              size = `Total reported direct emissions`)) +
  guides(size = FALSE) +
  labs(title = "All Carbon Point Sources in NY",
       subtitle = "(2023)")
```

Map only high purity sources in NY

```{r}
ggplot() +
  geom_sf(data = New_York, fill = NA) +
  geom_sf(data = data_NY_pure_carbon, 
          aes(color = `Industry Type (sectors)`,
          size = `Total reported direct emissions`)) +
  guides(size = FALSE) +
  labs(title = "High-Purity Carbon Point Sources in NY",
       subtitle = "(2023)")
```

##### Overlay Map

```{r}
ggplot() +
  geom_sf(data = New_York, fill = NA) +
  geom_sf(data = data_NY_pure_carbon, 
          aes(color = `Industry Type (sectors)`,
          size = `Total reported direct emissions`)) +
  guides(size = FALSE) +
  geom_sf(data = data_water_NY_above_mean, 
          aes(color = "Plant.Type")) +
  scale_fill_manual(name = "Industry Type", 
                    labels = c("Petroleum and Natural Gas Systems: High Purity Source", 
                               "High Volume Wastewater Facilities")) +
  labs(title = "Locations of suitable Wastewater and Carbon sources",
       subtitle = "For microalgae-cultivation in NY (2023)") 
```

Analyze futher with clustering analysis: determine top locations from clusters of water and carbon sources. 

Add in Leaflet Plot.

# Conclusions

Top locations for algae cultivation sites are noted by overlap of plotted points. 

distance matrix - points withing certain distance of each other
add image, add table, 


# References

